...

### 2023-09-16 manuscript/01_main.Rmd
revised methods

# Methodology

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially used patch of Central European temperate mixed forest (*Branišovský les*) close to České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure Norway spruce *Picea abies* plantations to mixed stands dominated by oaks (*Quercus* spp.). For more details on the species composition of the vegetation community, see *Supplementary Material section*.

@@

Vegetation features potentially relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was recorded according to whether asphalt roads ('Road') and forest-to-farmland transitions ('Edge') occurred within 150 m from the centre of the point or not ('Interior'). This was done using an aerial map (Mapy.cz). At each of the 32 points, all tree and shrub species within the 30 m radius were identified and counted, and the vertical strata to which they belonged noted ('Canopy' or 'Understorey'). Species identities were used to calculate species richness and diversity [Shannon's diversity index\; @shannon1948], while counts were used to calculate density per 100 m^2^. For other plants that classified as ground vegetation, only broad identifications were made (e.g., graminoids or mosses), and instead of counting them only the 'dominant' species (covering roughly $\ge$ 50% ground area) were identified. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch [@berdugo2018; @cornelissen2007]. *Section 1 in Supplementary Material* details the habitat characterisation methodology.

We also calculated the proportion of deciduous trees in the 'Canopy' as this has been shown to be a more important factor than measures of tree diversity [@berg1997]. Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10--15 m of the centre, which were then analysed using ImageJ software [@rasband1997] by converting to binary black-and-white images and calculating the percentage of cover across the whole image [see @forester2011\; *Figure S1*]. The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These canopy measurements were originally taken once in each sampling week, but most deciduous trees bore leaves even in the last week of sampling. Hence, the 13 weekly measurements of both canopy variables for each sampling point were instead averaged to provide more accurate measures of both, and these were used in the analyses. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period *(Section 2 in Supplementary Material)*.

@@

We surveyed bird 'abundance' by means of fixed-radius point counts, using detections as a proxy for abundance. While detections are not an accurate measure of true abundance, with many vegetation characteristics and behavioural traits affecting detectability of birds, it was sufficient for our purposes due to two reasons. First, we were not interested in estimating true population size, but rather in assessing relative habitat use using the proxy of relative abundance. Second, our sampling points did not vary greatly in terms of vegetation characteristics that hinder detectability, or in detectability of different bird communities. In fact, most birds have similarly low levels of vocal activity in the post-breeding period.

Point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate and separate two successive 'temporal units'. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point *(Table 1)*.

@@

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Each point count was associated with several variables *(see Table 2 in Supplementary Material)* such as Point ID, sampling day and week, observer, and other non-focal variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird's 'use' of the habitat.

All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as 'spuhs'. Prior to analysis, bird species were also classified based on feeding guild, which was determined using @gregory2007 (pt. Appendix 1) and Birds of the World [@billerman2020].

@@

To assess patterns of habitat selection by birds, we first excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. The measured habitat variables, without those excluded due to correlations, represented the pool of candidate predictors *(Table 2 in Supplementary Material)*. This set of plausible predictors was then used to build generalised linear mixed models (GLMMs) following an information-theoretic approach [@burnham1998]. That is, the Akaike Information Criterion [AIC\; @akaike1973; @akaike1974], and more specifically the corrected AIC or AIC~C~ [@hurvich1989], was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count, which was then used as the response variable. GLMMs were constructed using the *lme4* package [@bates2015], with the null model containing observer as a fixed effect (detections had *71:29* observer ratio) and sampling points and days as crossed random effects (model construction detailed later). Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function.

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficiently high number of overall detections, insectivores (rather, 'invertivores') and omnivores. Models were built separately for these two guilds, using the *glmmTMB* package [@brooks2017] to avoid convergence issues. The data summarised at guild level was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables were added. Following this, the other, focal variables were finally added *(see Table 1)*. Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package [@bolker2020], with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with Poisson error distribution and log link function fitted with the BOBYQA optimiser *(Table 2)*; the guild-level final models were a Poisson GLMM (log link) for invertivores and a negative binomial GLMM (log link) for omnivores.

For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Model diagnostics of all models were explored using the *DHARMa* package [@hartig2020] for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. All analyses were done using R version ***4.1.1*** [@rcoreteam2020]. Individual figures with multiple graphs were composed using the *patchwork* package for R [@pedersen2020a].

@@

## Conclusion

>>>>

# Methods

@@

The study was conducted over the course of 13 weeks, from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated patch of Central European temperate mixed forest (*Branišovský les*) close to České Budějovice, Czech Republic. The forest is located on the western edge of the city (48°58'55.672"N, 14°25'22.907"E), and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is very patchy owing to the long history of forestry regimes as well as its connection with rural and military Czech history. The patches range from pure Norway spruce plantations to mixed stands dominated by oaks (*Quercus* spp.). For more details on the species composition of the vegetation community, see *Supplementary Material section*.

@@

Vegetation features potentially relevant to birds in the post-breeding period were selected and measured to initially characterise the various habitat patches. The habitat class of each sampling point was defined according to whether asphalt roads ('Road') and forest-to-farmland transitions ('Edge') occurred within 150 m from the centre of the point or not ('Interior'). This was done using an aerial map (Mapy.cz). At each of the 32 points, all tree and shrub species within the 30 m radius were identified and counted, and the vertical strata to which they belonged noted ('Canopy' or 'Understorey'). Species identities were used to calculate species richness and diversity [Shannon's diversity index\; @shannon1948], while counts were used to calculate density per 100 m^2^. For other plants that classified as ground vegetation, only the 'dominant' species (covering roughly $\ge$ 50% ground area) at each point was recorded, for which identities were noted coarsely (e.g., graminoids or mosses) and counts were avoided. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality and disturbance of the forest patch [@berdugo2018; @cornelissen2007]. The habitat characterisation methodology is detailed in *Section 1 in Supplementary Material*.

We also calculated the proportion of deciduous trees in the 'Canopy', which is often more important than measures of tree diversity [@berg1997]. To measure canopy cover at a sampling point, canopy photographs were taken at three equidistant sub-points within 10--15 m of the centre, and then analysed using ImageJ software [@rasband1997] by converting to binary black-and-white images and calculating the percentage of cover across the whole image [see @forester2011\; *Figure S1*]. The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). These canopy measurements for each sampling point were taken once in each sampling week, as we originally expected canopy cover to vary as the season progressed. However, most deciduous trees bore leaves even in the last week of sampling, so the 13 weekly measurements of both canopy variables were instead averaged to provide single, more accurate measures of both. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period *(see Section 2 in Supplementary Material)*.

@@

Our objective was not to estimate true population size, but rather to assess relative habitat use, for which a proxy for bird abundance or density is suitable [see @vitz2007; @alatalo1981]. Moreover, vegetation characteristics that hinder detectability such as understorey density were generally low across our sampling points, and behaviourally, most birds have similarly low levels of vocal activity in the post-breeding period. Thus, there is little reason to expect major variation in detectability of bird communities across our sampling points, so although detections are not an accurate measure of true abundance, it served our purpose as a proxy for abundance.

We therefore surveyed bird abundance (using detections) by means of fixed-radius point counts. The point counts were done at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020), thus covering the post-breeding period from late summer till early autumn at a fine-grained temporal resolution. Sampling was done on 2--3 days in the middle of the calendar week while the remaining days served to delineate two successive 'temporal units'. This enabled meaningful comparisons of bird abundances across weeks, as each week would be sufficiently independent of the others and would thus represent a distinct time point *(Table 1)*.

@@

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Each sampling event (point count) was associated with several variables *(see Table 2 in Supplementary Material)* such as Point ID, sampling day and week, observer, and other non-focal variables such as weather, visibility, and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen during a period of 20 minutes, divided into four intervals of five minutes each. Moreover, complementary information for each detection like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and $\gt$ 50 m. In addition, birds were observed for whether or not they entered the 30 m radius (i.e., entered the sampling point) at any time during the count, signifying the bird's 'use' of the habitat.

All birds were identified to species level when possible, but some closely related cryptics such as *Phylloscopus* spp., *Certhia* spp., *Regulus* spp. and *Dendrocopos* spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as 'spuhs'. Prior to analysis, bird species were also classified based on feeding guild, which was determined using Gregory et al. [-@gregory2007\; Appendix 1] and Birds of the World [@billerman2020].

@@

To assess patterns of habitat selection by birds, we first excluded from our dataset birds not identified to genus level, those detected flying over the forest, and those that did not enter the 30 m radius. The list of measured habitat variables, excluding those removed due to correlations, represented the pool of candidate predictors *(Table 2 in Supplementary Material)*. This set of plausible predictors was then used to build generalised linear mixed models (GLMMs) following an information-theoretic approach [@burnham1998]. That is, the Akaike Information Criterion [AIC\; @akaike1974; @akaike1973], and more specifically the corrected AIC or AIC~C~ [@hurvich1989], was used to judge model fit.

Analyses were conducted at two main levels of response. At the highest level, we looked for patterns of habitat selection in the whole bird community, by summing all the individual species detections per point count to obtain one 'bird abundance' value for each count as the response variable. The null model was constructed with observer as a fixed effect (detections had *71:29* observer ratio), and sampling points (point ID) and days as crossed random effects (model construction detailed later). Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day in order to account for the random variability in each unique day. Residuals were modelled as coming from a Poisson distribution with a log link function.

At the second level, we investigated whether foraging guilds differed in their habitat selection patterns. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had sufficient overall detections for analysis, invertivores and omnivores, and they were modelled separately. This guild-level data was insufficient for utilisation of two crossed random effects (as was done for the community-level models), so the random effect of sampling day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates and so was not considered to be an issue. The null models had the same fixed effect as in the first case.

GLMMs were fitted using the *glmmTMB* package [@brooks2017], which performed better than the *lme4* [@bates2015] alternative when dealing with random effects having very low variance (point ID in the community-level models), and also avoided convergence issues in the guild-level models. Models were constructed following a null model approach, i.e., starting from a null model then building up using stepwise selection of predictors. First, the week variable which had a visible effect (abundances were higher towards the end of the study period; noticed even during the bird counts), and other non-focal variables *(see Section 2.3)* were added. Following this, the focal variables were added *(see Table 1)*.

Stepwise selection was done by comparing AIC~C~ values of all the models using the AICctab() function from the *bbmle* package [@bolker2020], with an information threshold of 2 dAIC~C~ per df. If a variable was deemed important, its interaction with 'Week' was next tested (i.e., whether or not habitat selection criteria changed with the progress of the season). Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was selected. At the community level, the final model was a GLMM with negative binomial error distribution and log link function *(Table 2)*; the guild-level final models were a Poisson GLMM (log link) for invertivores and a negative binomial GLMM (log link) for omnivores.

For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. Model diagnostics of all models were explored using the *DHARMa* package [@hartig2020] for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. We used the bootMer() function in *lme4* in order to test the significance of the fitted model results, by obtaining mean and CI estimates using semi-parametric bootstrapping. The bootstrapping was done with 1000 simulations and with parallel core processing [with the help of packages *parallel*, @rcoreteam2020; and *VGAM*, @yee2010]. Figures comprising multiple graphs were composed using the *patchwork* package for R [@pedersen2020a]. All analyses were done using `r R.version.string` [@rcoreteam2020].

@@

# Conclusion
